---
title: "CCSM4 Europe"
author: "Nick Gauthier"
date: "September 18, 2015"
output: html_document
---

In this notebook, we estimate mid-Holocene monthly mean precipitation and minimum and maximum precipitation in the Mediterranean and southern Europe by statistically downscaling General Circulation Model (GCM) paleoclimate simulation. We first calibrate a Genralized Additive Model (GAM) using observed WorldClim climatologies, DEM-derived topographic variables, and simulations of present day climates using NCAR's CESM/CCSM4 model. Then we apply this model to CCSM4 simulations of the mid-Holocene PMIP3 experiments.
 
 ##Preparation
 ###Setup
 
 First load the necessary R packages.

``` {r message= F, warning = F}
library(raster)     # functions for managing and processing raster data
library(mgcv)       # functions to fit and analyze GAMs
library(dismo)      # functions to sample points weighted by latitude
library(magrittr)   # piping functions for code readability
library(rasterVis)  # functions for visualizing raster maps
library(ggplot2)    # functions for plotting
library(reshape2)
```

### Observed Climate

Import the observed present-day climatologies which we'll use to calibrate the model. We use WorldClim data at 5min resolution here, which can be easily downloaded using the *getData()* function in the *raster* package. 

```{r, message=F}
get.wc <- function(var){
     raster::getData('worldclim', var=var, res = 5, download = T) %>% 
     crop(extent(-20, 50, 25, 58)) %>%     
     projectRaster(crs = '+proj=laea') %>%
     crop(extent(c(-999742, 3770408, 3773858, 5468408))) %>%
     set_names(month.name)
}

obs.prc <- get.wc('prec')
obs.tmx <- get.wc('tmax') %>% divide_by(10) # raw data in degrees Celsius * 10
obs.tmn <- get.wc('tmin') %>% divide_by(10) 
```

```{r echo = F}
levelplot(obs.prc[[c(1, 7)]], par.settings = rasterTheme(region=brewer.pal('Blues', n=9)))

levelplot(brick(c(obs.tmn[[c(1,7)]], obs.tmx[[c(1,7)]])), par.settings = BuRdTheme)

ggplot(melt(values(obs.prc)), aes(x=Var2, y=value, group=Var1)) + 
     geom_line(alpha = .01) +
     theme_classic()

ggplot(melt(values(obs.tmx)), aes(x=Var2, y=value, group=Var1)) + 
     geom_line(alpha = .01) +
     theme_classic()
#### do these pots for the predicted va;ues as well!
```


### GCM Predictors
```{r}
import.trace <- function(dir, var, sim = 'mh6k', level = 1){
     gcm.in <- brick(dir, var = var, level = level) 
     
     if(sim == 'control'){gcm.in <- gcm.in[[1200:1799]]}
     
     gcm.in <- stackApply(gcm.in, indices = 1:12, fun = mean)
     extent(gcm.in) <- extent(0, 360, -90, 90)
     
     gcm.in %>%    
     rotate %>%
     projectRaster(obs.prc) %>%
     mask(obs.prc)
}

control.prc <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.PRECT.185001-200512.nc', 
                            var = "PRECT", 
                            sim = 'control') %>% multiply_by(2629743830)

control.t <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.TREFHT.185001-200512.nc', 
                            var = "TREFHT", 
                            sim = 'control') %>% subtract(273.15)

control.tmx <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.TREFMXAV.185001-200512.nc', 
                            var = "TREFMXAV", 
                            sim = 'control') %>% subtract(273.15)

control.tmn <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.TREFMNAV.185001-200512.nc', 
                            var = "TREFMNAV", 
                            sim = 'control') %>% subtract(273.15)

control.psl <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.PSL.185001-200512.nc', 
                            var = "PSL", 
                            sim = 'control') %>% divide_by(1000)

control.rh <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.RHREFHT.185001-200512.nc', 
                            var = "RHREFHT", 
                            sim = 'control')

control.q <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.QREFHT.185001-200512.nc', 
                           var = "QREFHT", 
                           sim = 'control')

control.u <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.U.185001-200512.nc', 
                          var = "U", 
                          sim = 'control',
                          level = 26)

control.v <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.V.185001-200512.nc', 
                          var = "V", 
                          sim = 'control',
                          level = 26)

control.z <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.20th.track1.1deg.005.cam2.h0.Z3.185001-200512.nc', 
                          var = "Z3", 
                          sim = 'control',
                          level = 23) # ~850 geopotential


#plot(brick(control.prc[[1]], control.psl[[1]], control.q[[1]], control.rh[[1]], control.tmn[[1]], control.tmx[[1]], control.z[[1]]))

#pairs(brick(control.prc[[1]], control.psl[[1]], control.q[[1]], control.rh[[1]], control.tmn[[1]], control.tmx[[1]], control.z[[1]], obs.prc[[1]], obs.tmn[[1]],obs.tmx[[1]]))
```

### Topographic Predictors
Generate some topographic predictors from a DEM.

Start with a DEM.
```{r}
elev <- raster('alt_5m_bil/alt.bil') %>% projectRaster(obs.prc)

levelplot(elev, par.settings = GrTheme, margin = F)
```

Calculate the euclidean distance to the ocean, accounting for latitude effects.
```{r}
dco <- raster('alt_5m_bil/alt.bil') %>% # import WorldClim 5m dem
    crop(extent(-20, 50, 25, 58)) %>%  # start with a wider region to get accurate distances
    reclassify(c(-Inf, Inf, NA, NA, NA, 1)) %>% # reverse NA and non-NA cells
    distance(doEdge = T) %>% # calculate the distances
    projectRaster(obs.prc) %>%
    mask(elev) %>% # mask out ocean cells  
    divide_by(1000) # convert to km

levelplot(dco, margin = F)
```

Calculate the distance to the ocean in the direction along incoming air mass paths, a proxy for advective continentality.

```{r}
dir <- (270 - overlay(control.v,  control.u, fun = atan2) * (180 / pi)) %% 360 # in degrees into which the wind is blowing is 90-, from which is 270 -
#http://stackoverflow.com/questions/21484558/how-to-calculate-wind-direction-from-u-and-v-wind-components-in-r

streamplot(stack(control.u[[1]], control.v[[1]]), isField = 'dXY')
streamplot(stack(control.u[[7]], control.v[[7]]), isField = 'dXY')


##convert to r-compatible flow direction
flow.dir <- reclassify(dir, c(-1, 22.5, 64, 
                  22.5, 67.5, 128, 
                  67.5, 112.5, 1,
                  112.5, 157.5, 2,
                  157.5, 202.5, 4,
                  202.5, 247.5, 8,
                  247.5, 292.5, 16,
                  292.5, 337.5, 32,
                  337.5, 361, 64))
```

The following chunk of code calculates ACO, the calculations take a long time (~8 hrs on a macbook) and are not run here. Instead, a preprepared ACO dataset is imported in the next chunk for later analyses. If you need to calculate aco, copy and paste this code into your R console. 
```{r eval = F}
indices <- which(!is.na(values(elev)))
pb <- txtProgressBar(min = 1, max = 12, style = 3) 
aco <- flow.dir
for(i in 1:12){
     direction.map <- flow.dir[[i]]
     test <- mclapply(indices, mc.cores = 8, function(x){
          flowPath(direction.map, x) %>% length})
     values(aco[[i]])[indices] <- test %>% unlist
     setTxtProgressBar(pb, i)
     }
close(pb)
```

```{r}
aco <- brick('aco') %>% projectRaster(obs.prc)
levelplot(aco)
```

Lets think about some more topographic predictors
one option is sin and cos of the aspect in radians, which gives you an idea of the relative e-w and n-s exposures, respectively

```{r}
#ew.exposure <- elev %>% terrain(opt='aspect') %>% sin
#ns.exposure <- elev %>% terrain(opt='aspect') %>% cos
aspect <- elev %>% terrain(opt='aspect', unit = 'degrees')
slope <- elev %>% terrain(opt = 'slope', unit = 'degrees')
```

```{r, eval = F}
# need to check out that Baopo Fu refernce to confirm this

vel <- sqrt(control.u ^ 2 + control.v ^ 2)

#distance betwwn wind angle and slope orientation
delta <- abs(dir - aspect)
values(delta) <- ifelse(values(delta) > 180, 360 - values(delta), values(delta))
plot(delta[[1]])     


vg <- sin(slope * 2 * pi / 180) * cos(delta * pi / 180) * vel * .5
plot(vg[[1]]*control.prc[[1]])
```
Latitude

```{r}
lat <- elev %>%
     coordinates %>%
     extract(,2) %>% 
     setValues(elev, .)

lon <- elev %>%
     coordinates %>%
     extract(,1) %>% 
     setValues(elev, .)
```


monthly
```{r}
month <- obs.tmx %>% setValues(rep(1:12, each = 126555))
```

Put everything together.

```{r}
cal.vars <- sapply(1:12, function(x){ 
  brick(obs.tmx[[x]], obs.tmn[[x]], obs.prc[[x]], control.tmx[[x]], control.tmn[[x]], control.q[[x]], control.rh[[x]], control.prc[[x]], control.psl[[x]], control.z[[x]], elev, dco, vg[[x]],lat,lon, month[[x]]) %>%
    setNames(c('obs.tmx', 'obs.tmn', 'obs.prc','TREFMXAV', 'TREFMNAV', 'Q', 'RH', 'PRC', 'PSL', 'Z3', 'elev','dco', 'vg', 'lat','lon', 'month'))
  })
rm(control.tmx, control.tmn, control.q, control.rh,control.prc,control.psl,control.z, dco,aco,ew.exposure,exposure,aspect,season,lat,lon)
```

Sample the variables at random points, weighting for area distortions due to latitude
```{r}
cal.data <- lapply(cal.vars, function(x) (raster::extract(x, randomPoints(elev, 20000)) %>% data.frame)) %>% do.call(rbind, .)

cal.data$month <- cal.data$month %>% factor()


cal.data <- lapply(cal.vars, values) %>% do.call(rbind, .) %>% na.omit %>% data.frame

pairs(cal.data)
qplot(cal.data$obs.prc,cal.data$PRC, alpha = I(.1))
# good correlation, lets start with a gam for these two vars and work from there
```

## Fit the model

let's start with temperature
```{r}

fit.tmn <- gam(obs.tmn ~ s(TREFMNAV, bs = 'cr') +
                         s(Z3, bs = 'cr') +
                         s(elev, bs = 'cr') +
                         s(dco, bs = 'cr') +
                         s(month, bs = 're'),
                method = 'REML', data = cal.data)


fit.tmn
summary(fit.tmn)

#93% deviance explained
#75% with just temp elev and dco and latlong
gam.check(fit.tmn)
#looks good, normal rerrors, overpredicts in top quantile

plot(fit.tmn, shade=T, seWithMean = T, pages = 1)

concurvity(fit.tmn)
concurvity(fit.tmn,full = F)$observed

library(GGally)
ggpairs(cal.data, c(5:10), alpha = .1)
pairs(cal.data[,5:10], pch = '.')

month <- 1
fit.error <- predict(cal.vars[[month]], fit.tmn, type = 'response') - obs.tmn[[month]]
levelplot(fit.error,  par.settings = RdBuTheme, at = seq(-10,10, 2), contour = F, margin = F)
levelplot(brick(predict(cal.vars[[month]], fit.tmn, type = 'response'), obs.tmn[[month]]))

# check for autocorrelatio in the residuals
Moran(fit.error)
library(usdm)
Variogram(fit.error, size = 1000) %>% plot
# looks quadratic?
```

```{r}
fit.tmx <- gam(obs.tmx ~ s(TREFMXAV, bs = 'cr') +
                         s(Z3, bs = 'cr') +
                         s(elev, bs = 'cr') +
                         s(dco, bs = 'cr') +
                         s(month, bs = 're'),
                method = 'REML', data = cal.data)

summary(fit.tmx)
gam.check(fit.tmn)
plot(fit.tmx, shade = T, seWithMean = T, pages = 1)

fit.tmx.noseason <- gam(obs.tmx ~ s(TREFMXAV, bs = 'cr') +
                         s(Z3, bs = 'cr') +
                         s(elev, bs = 'cr') +
                         s(dco, bs = 'cr'),
                        method = 'REML', data = cal.data)

fit.tmx.noz <- gam(obs.tmx ~ s(TREFMXAV, bs = 'cr') +
                         s(elev, bs = 'cr') +
                         s(dco, bs = 'cr') +
                         s(month, bs = 're'),
                method = 'REML', data = cal.data)
```

```{r}
fit.prc.occurrence <- gam(factor(obs.prc >= 1) ~ s(PRC),
                    family = binomial, method = 'REML', data = cal.data)
#fitting with other vars shrinks everything to zero leaving only prc

summary(fit.prc.occurrence)
month <- 8
levelplot(predict(cal.vars[[month]], fit.prc.occurrence, type = 'response'))
plot(predict(cal.vars[[month]], fit.prc.occurrence, type = 'response') < .6)

fit.prc.occurrence %>% summary
plot(fit.prc.occurrence, seWithMean = T, shade = T, pages = 1, scale = 0)

fit.prc <- bam(obs.prc ~ s(PRC, bs = 'cr') +
                         s(PSL, bs = 'cr') +
                         s(Q, bs = 'cr') + 
                         s(vg, bs = 'cr') +
                         s(Z3, bs = 'cr') +
                         s(elev, bs = 'cr') +
                         s(dco, bs = 'cr') +
                         s(month, bs = 're'),
               family = Gamma(link = 'log'), method = 'REML', data = cal.data[cal.data$obs.prc>=1,])

summary(fit.prc)
fit.prc
concurvity(fit.prc)
gam.check(fit.prc)
plot(fit.prc, shade = T, seWithMean = T, pages = 1)
plot(fit.prc, residuals = T)



month <- 3
fit.error <- predict(cal.vars[[month]], fit.prc, type = 'response') - obs.prc[[month]]
levelplot(fit.error,  par.settings = RdBuTheme, at = seq(-175, 175, 25),contour = F, margin = F)
levelplot(fit.error/obs.prc[[month]] * 100,  par.settings = RdBuTheme, at = seq(-100, 100, 25),contour = F, margin = F)

levelplot(brick(predict(cal.vars[[month]], fit.prc, type = 'response'), obs.prc[[month]]))

```


TEST models on out of sample domain

```{r}
pred.vars <- sapply(1:12, function(x){ 
  brick(mh6k.tmx[[x]], mh6k.tmn[[x]], mh6k.prc[[x]], mh6k.psl[[x]], mh6k.z[[x]], elev, dco,lat,lon) %>%
    setNames(c('TREFMXAV', 'TREFMNAV', 'PRC', 'PSL', 'Z3', 'elev','dco', 'lat','lon'))
  })
```


```{r}
tmn.recon <- lapply(1:12, function(x){predict(pred.vars[[x]], fit.tmn, type = 'response')}) %>% brick
tmx.recon <- lapply(1:12, function(x){predict(pred.vars[[x]], fit.tmx, type = 'response')}) %>% brick
prc.occurrence.recon <- lapply(1:12, function(x){predict(pred.vars[[x]], fit.prc.occurrence, type = 'response')}) %>% brick
prc.recon <- lapply(1:12, function(x){predict(pred.vars[[x]], fit.prc, type = 'response')}) %>% brick
prc.recon <- mask(prc.recon, prc.occurrence.recon>.7, maskvalue=0, update.value=0)


levelplot(prc.occurrence.recon)
levelplot(prc.recon - obs.prc, par.settings = RdBuTheme, at = seq(-200,200,20))
levelplot(tmn.recon - obs.tmn, par.settings = BuRdTheme, at = seq(-15,15,2))
levelplot(tmx.recon - obs.tmx, par.settings = BuRdTheme, at = seq(-20,20,2))

writeRaster(tmn.recon, 'tmn_mh6k.tif', format = 'GTiff', options="INTERLEAVE=BAND", overwrite = T)
writeRaster(tmx.recon, 'tmx_mh6k.tif', format = 'GTiff', options="INTERLEAVE=BAND", overwrite = T)
writeRaster(prc.recon, 'prc_mh6k.tif', format = 'GTiff', options="INTERLEAVE=BAND", overwrite = T)

```

 
```{r}
mh6k.tmx <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.mh6ka.1deg.003.cam2.h0.TREFMXAV.080101-131612.nc', 
                         var = "TREFMXAV") %>% subtract(273.15)

mh6k.tmn <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.mh6ka.1deg.003.cam2.h0.TREFMNAV.080101-131612.nc', 
                         var = "TREFMNAV") %>% subtract(273.15)

mh6k.z <- import.trace('/Volumes/Time Machine Backups/b40.mh6ka.1deg.003.cam2.h0.Z3.080101-131612.nc', 
                         var = "Z3", level = 23)

mh6k.psl <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.mh6ka.1deg.003.cam2.h0.PSL.080101-131612.nc', 
                         var = "PSL") %>% divide_by(1000)

mh6k.prc <- import.trace('~/Google Drive/Climate Data/Models/CCSM4/b40.mh6ka.1deg.003.cam2.h0.PRECT.080101-131612.nc', 
                         var = "PRECT") %>% multiply_by(2629743830)

mh6k.q <- import.trace('~/Google Drive/b40.mh6ka.1deg.003.cam2.h0.QREFHT.080101-131612.nc', 
                           var = "QREFHT")

mh6k.u <- import.trace('/Volumes/Time Machine Backups/b40.mh6ka.1deg.003.cam2.h0.U.080101-131612.nc', 
                          var = "U",
                          level = 26)

mh6k.v <- import.trace('/Volumes/Time Machine Backups/b40.mh6ka.1deg.003.cam2.h0.V.080101-131612.nc', 
                          var = "V",
                          level = 26)

mh6k.fsds <- import.trace('~/Downloads/b40.mh6ka.1deg.003.cam2.h0.FSDS.080101-131612.nc', var = 'FSDS')
plot(mh6k.fsds)
writeRaster(mh6k.fsds, 'fsds_mh6k.tif', format = 'GTiff', options="INTERLEAVE=BAND", overwrite = T)


writeRaster(mh6k.z,'mh6kz')
writeRaster(mh6k.v, 'mh6kv')
writeRaster(mh6k.u, 'mh6ku')

dir.6k <- (270 - overlay(mh6k.v,  mh6k.u, fun = atan2) * (180 / pi)) %% 360
vel.6k <- sqrt(mh6k.u ^ 2 + mh6k.v ^ 2)

#distance betwwn wind angle and slope orientation
delta.6k <- abs(dir.6k - aspect)
values(delta.6k) <- ifelse(values(delta.6k) > 180, 360 - values(delta.6k), values(delta.6k))
vg.6k <- sin(slope * 2 * pi / 180) * cos(delta.6k * pi / 180) * vel.6k * .5

pred.vars <- sapply(1:12, function(x){ 
  brick(mh6k.tmx[[x]], mh6k.tmn[[x]], mh6k.prc[[x]], mh6k.psl[[x]], mh6k.q[[x]], mh6k.z[[x]], elev, dco, vg.6k[[x]], month[[x]]) %>%
    setNames(c('TREFMXAV', 'TREFMNAV', 'PRC', 'PSL', 'Q','Z3', 'elev','dco', 'vg','month'))
  })
```

Rain Days
```{r}

mh6k.prect.daily <- brick('~/Downloads/b40.mh6ka.1deg.003M.cam2.h1.PRECC.12690101-13001231.nc') + 
     brick('~/Downloads/b40.mh6ka.1deg.003M.cam2.h1.PRECL.12690101-13001231.nc') %>% 
     multiply_by(1000 * 60 * 60 * 24)

day.to.month <- c(rep(1,31),rep(2,28),rep(3,31),rep(4,30),rep(5,31),rep(6,30),rep(7,31),rep(8,31),rep(9,30),rep(10,31),rep(11,30),rep(12,31))

for(i in seq(12,372,12)){
     day.to.month <- c(day.to.month,rep(1+i,31),rep(2+i,28),rep(3+i,31),rep(4+i,30),rep(5+i,31),rep(6+i,30),rep(7+i,31),rep(8+i,31),rep(9+i,30),rep(10+i,31),rep(11+i,30),rep(12+i,31))     
}

mh6k.wet <- (mh6k.prect.daily > 2) %>%
     stackApply(indices = day.to.month, fun = sum) %>%
     stackApply(indices = 1:12, fun = mean)
     
extent(mh6k.wet) <- extent(0, 360, -90, 90)
     
mh6k.wet <- mh6k.wet %>%    
     rotate %>%
     projectRaster(obs.prc) %>%
     mask(obs.prc)



levelplot(mh6k.wet)
writeRaster(mh6k.wet, 'wet_mh6k.tif',format = 'GTiff', options="INTERLEAVE=BAND", overwrite = T)
```

