---
title: 'Supplementary Information: Paleoclimate Modeling'
author: "Nicolas Gauthier"
date: 'Last Updated: `r Sys.Date()`'
link-citations: yes
mainfont: Times New Roman
monofont: Ubuntu Mono
geometry: margin=1in
fontsize: 12pt
header-includes:
- \usepackage{marginnote}
output:
  pdf_document:
    fig_height: 5
    fig_width: 10
    highlight: pygments
    latex_engine: lualatex
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Introduction

We'll be comparing paleoclimate model estimates of temperature and precipitation over three points in the west Mediterranean to global paleoclimate proxies.

## Setup
Load all the packages we'll need for this analysis. 

`r margin_note("You'll need to have the netCDF libraries already installed on your system for ncdf4 to work.")`
```{r libraries, message=F}
library(ncdf4) # import GCM data
library(rgdal) # read GCM data
library(raster) # process GCM data
library(rasterVis) # plotting GCM data
library(tidyverse) # data management and plotting
library(magrittr) # pipes for code readability
library(EMD) # calculate trends in the data
library(dismo) # for latitudinally weighted samples
library(mgcv) # fit GAM for downscaling
```


# Climate Model

## Sample Locations
```{r map, echo = F, fig.margin = T, message = F, cache = T, fig.cap = 'Locations of 3 sample points.'}
library(ggmap)

samp.pts <- matrix(c(0,40, 4,44, 12, 46, 14, 43), ncol = 2, byrow = T) %>% as_data_frame

get_map(location = c(left = -10, bottom = 35, right = 20, top = 50), 
        maptype = 'terrain-background', source = 'stamen') %>% 
  ggmap + labs(x = 'Longitude', y = 'Latitude') + geom_point(data = samp.pts, aes(x = V1, y = V2), size = 3, color = 'red')
```

Create a matrix with the coordinates for the three locations of interest in the west Mediterranean.
```{r samppoints}
samp.pts <- matrix(c(0, 40, 4, 44, 12, 46, 14, 43), 
                   ncol = 2, byrow = T)

#12,43
```

## TraCE-21k

```{r trace_map, echo=FALSE, fig.show = 'hold', fig.cap = 'TraCE21-k global precipitation and temperature', fig.margin=TRUE, message=F}
brick('trace.01-36.22000BP.cam2.PRECT.22000BP_decavg_400BCE.nc') %>%
  extract2(1) %>%
  multiply_by(3.154e+10) %>%
  levelplot(margin = F, par.settings = PuOrTheme())

brick('trace.01-36.22000BP.cam2.TREFHT.22000BP_decavg_400BCE.nc') %>%
  extract2(1) %>%
  subtract(273.15) %>%
  levelplot(margin = F, par.settings = BuRdTheme())
```

First, import data from the TraCE-21k paleoclimate simulation. Then extract temperature and precipitation values at three locations in the west Mediterranean. Use the *brick* function from **raster** to import decadal averages from the simulation. Put the coordinates for the three locations in a matrix, and use that matrix to and **raster's** *extract* function to get the values from the climate model brick. Convert the precipitation values to mm/year and temperature values to degrees Celsius. Finally, name the columns for each region appropriately. 

Now pull all the TraCE data into one data frame, with one row per year, and one column per variable/location combination. First *rbind* the two sets of TraCE data and *transpose* the results, turning the 6 rows into 6 columns. Add a column for the Year (in ka BP), and use to select only the entries earlier than 6,000 BP.

```{r}
trace.dat <- rbind(
  brick('trace.01-36.22000BP.cam2.TREFHT.22000BP_decavg_400BCE.nc') %>%
    raster::extract(samp.pts) %>% 
    subtract(273.15), # convert from kelvin to C
  brick('trace.01-36.22000BP.cam2.PRECT.22000BP_decavg_400BCE.nc') %>%
    raster::extract(samp.pts) %>% # extract data at these coordinates
    multiply_by(3.154e+10)) %>% # convert to mm/year
  t %>% # transpose
  as.data.frame %>%
  set_colnames(c('tmp,Southwest', 'tmp,North Central', 'tmp,Northeast', 'tmp,Southeast',
                 'prc,Southwest', 'prc,North Central', 'prc,Northeast', 'prc,Southeast')) %>%
  rownames_to_column('Year') %>%
  mutate(Year = as.numeric(substring(Year, 3))) %>%
  filter(Year > 6) # get all the decades up to 6ka BP
```



## Trend Analysis

Let's use the **EMD** package to calculate actual trend lines using the empirical mode decomposition approach.
```{r, echo=FALSE, fig.cap="Empirical mode decomposition", fig.height = 10, fig.margin=TRUE, message=F, cache = F}
emd.test <- emd(trace.dat[,2])

par(mfrow=c(emd.test$nimf+1, 1), mar=c(2,1,2,1))
rangeimf <- range(emd.test$imf)
for(i in 1:emd.test$nimf) {
  plot(1:1600, emd.test$imf[,i], type="l", xlab="", ylab="", ylim=rangeimf,
       main=paste(i, "-th IMF", sep="")); abline(h=0)
}
```

Now organize the temperature and precipitation data to make plotting easier using functions from **tidyr**.
`r margin_note("Replace the variable names to make facet naming easier too.")`
```{r cache = F}
trace.plot <- trace.dat %>% 
  gather(key, value, - Year) %>%
  separate(key, c('Variable', 'Region'), ',') %>%
  mutate(Region = factor(Region, levels = c('Southwest', 'North Central', 'Northeast', 'Southeast')),    # turn region into factor so it is ordered correctly
         Variable = ifelse(
    Variable == 'tmp', 'Temperature (°C)', 'Precipitation (mm)'))

emd.res <- function(x) emd(x, boundary = 'wave')$residue
trace.emd <- trace.dat %>%
  mutate_at(vars(-Year), emd.res) %>%
  gather(key, value, - Year) %>%
  separate(key, c('Variable', 'Region'), ',') %>%
  mutate(Region = factor(Region, levels = c('Southwest', 'North Central', 'Northeast', 'Southeast')),
         Variable = ifelse(
    Variable == 'tmp', 'Temperature (°C)', 'Precipitation (mm)'))

#trace.plot %>% mutate(Raw = value, Trend = trace.emd[,4], value = NULL) %>% write_csv('~/Downloads/trace_plot_data.csv')
```

## Plotting
Plot everything with **ggplot2**.
```{r bwplot, echo=FALSE, fig.cap="Black and white version", fig.width = 16.9, fig.height = 8.61, fig.margin=TRUE, message=F, cache = F}
ggplot(data = trace.plot, aes(x = Year, y = value)) +
  facet_grid(Variable ~ Region, switch = 'y', scale = 'free_y') +
  geom_vline(xintercept = c(22, 19, 14, 10, 6), lty = 2) +
  geom_point(color = 'grey', alpha = .3) +
  geom_line(data = trace.emd, size = 1.2, color = "black", alpha = .8) +
  scale_x_reverse(breaks = seq(6,22,4)) +
  guides(color = "none") +
  labs(x = '1,000 Years BP', y = '') +
  theme_bw(base_size = 20) +
  theme(strip.background  = element_blank())
```

```{r mainplot, fig.fullwidth = TRUE, fig.width = 16.9, fig.height = 8.61}
ggplot(data = trace.plot, aes(x = Year, y = value)) +
  facet_grid(Variable ~ Region, switch = 'y', scale = 'free_y') +
  geom_vline(xintercept = c(22, 19, 14, 10, 6), lty = 2) +
  geom_point(aes(color = Variable), alpha = .3) +
  geom_line(data = trace.emd, size = 1.2, color = "black", alpha = .8) +
  scale_x_reverse(breaks = seq(6,22,4)) +
  labs(x = '1,000 Years BP', y = '') +
  guides(color = "none") +
  theme_bw(base_size = 20) +
  theme(strip.background  = element_blank())
```

Calculate the detrended variances.
```{r fig.fullwidth = T}
emd.dat <- trace.dat %>%
  mutate_at(vars(-Year), emd.res)

(trace.dat - emd.dat) %>%
  select(-Year) %>%
  cbind(Year = trace.dat$Year, .) %>%
  mutate(Period = cut(Year, c(22, 19, 14, 10, 6))) %>%
  group_by(Period) %>%
  summarise_each(funs(var)) %>%
  select(-Year) %>%
    print(width = Inf)
```

How do these compare to the overall regional variance?
```{r}
bbox <- extent(c(-10, 20, 35, 47))

trace.reg.avg <- rbind(
  brick('trace.01-36.22000BP.cam2.TREFHT.22000BP_decavg_400BCE.nc') %>%
    raster::extract(bbox, fun = mean) %>% 
    subtract(273.15), # convert from kelvin to C
  brick('trace.01-36.22000BP.cam2.PRECT.22000BP_decavg_400BCE.nc') %>%
    raster::extract(bbox, fun = mean) %>%
    multiply_by(3.154e+10)) %>% # convert to mm/year
  t %>% # transpose
  as.data.frame %>%
  set_colnames(c('tmp,StudyArea', 'prc,StudyArea')) %>%
  rownames_to_column('Year') %>%
  mutate(Year = as.numeric(substring(Year, 3))) %>%
  filter(Year > 6) 

emd.reg.avg <- trace.reg.avg %>%
  mutate_at(vars(-Year), emd.res)

(trace.dat - emd.dat) %>%
  select(-Year) %>%
  cbind(Year = trace.dat$Year, .) %>%
  mutate(Period = cut(Year, c(22, 19, 14, 10, 6))) %>%
  group_by(Period) %>%
  summarise_each(funs(var)) %>%
  select(-Year) %>%
  subtract(((trace.reg.avg - emd.reg.avg) %>%
  magrittr::extract(c(2,2,2,2,3,3,3,3)) %>%
  cbind(Year = trace.dat$Year, .) %>%
  mutate(Period = cut(Year, c(22, 19, 14, 10, 6))) %>%
  group_by(Period) %>%
  summarise_each(funs(var)) %>%
  select(-Year))) %>%
    print(width = Inf)
```



## Spatial patterns



# Data Preprocessing

## Observations
```{r}
bbox <- extent(c(-10, 45, 30, 50))
  
tmean.obs <- list.files('~/gdrive/Data/MOD11C3v5.0-CHIRPSv2.0_MONTHLY_03m/meantemp', full.names = T) %>% 
    stack %>%
    set_names(month.name) %>%
    crop(bbox)
tmean.obs[tmean.obs == -9999] <- NA

p.obs <- list.files('~/gdrive/Data/MOD11C3v5.0-CHIRPSv2.0_MONTHLY_03m/precip', full.names = T) %>% 
    stack %>%
    set_names(month.name) %>%
    crop(bbox)
p.obs[p.obs == -9999] <- NA
```


## Topographic Variables
```{r}
elev <- raster('~/gdrive/Data/SRTM_1km.tif') %>% projectRaster(p.obs) %>% mask(p.obs)
```


```{r cache = T}
dco <- elev %>%
  reclassify(c(-Inf, Inf, NA, NA, NA, 1)) %>% # reverse NA and non-NA cells
  distance(doEdge = T) %>% # calculate the distances
  mask(elev) %>% # mask out ocean cells  
  divide_by(1000) # convert to km
```



## Reanalysis
Import and preprocess of ECMWF-interim reanalysis data, monthly means of daily means, 1979-2010
```{r eval = FALSE}
processECMWF <- function(file, var){
  brick(paste0('~/gdrive/Data/', file), varname = var) %>% 
      stackApply(indices = 1:12, fun = mean) %>%
      rotate %>%
      set_names(month.name) %>% projectRaster(eu.p) %>% mask(eu.p)
}

tcw <- processECMWF('ecmwf_surface.nc', 'tcw')
msl <- processECMWF('ecmwf_surface.nc', 'msl')
t2m <- processECMWF('ecmwf_surface.nc', 't2m')
lsp <- processECMWF('ECMWF Precip.nc', 'lsp')
cp <- processECMWF('ECMWF Precip.nc', 'cp')
```

```{r, include=FALSE}
# load preloaded files at first for convience, should change
msl <- brick('~/gdrive/Data/ECMWF/eu/msl_ecmwf_eu.tif')
t2m <- brick('~/gdrive/Data/ECMWF/eu/t2m_ecmwf_eu.tif')
tcw <- brick('~/gdrive/Data/ECMWF/eu/tcw_ecmwf_eu.tif')
cp <- brick('~/gdrive/Data/ECMWF/eu/cp_ecmwf_eu.tif')
lsp <- brick('~/gdrive/Data/ECMWF/eu/lsp_ecmwf_eu.tif')
```


Put all the predictor and response variables together, month by month.
```{r}
cal.vars <- sapply(1:12, function(x){ 
  brick(tmean.obs[[x]], p.obs[[x]], msl[[x]], t2m[[x]], tcw[[x]], lsp[[x]], cp[[x]], elev, dco) %>%
    setNames(c('tmean.obs', 'p.obs', 'msl', 't2m', 'tcw', 'lsp', 'cp','elev','dco'))
})
```

Sample the variables at random points, weighting for latitude
```{r}
cal.data <- lapply(cal.vars, function(x) (raster::extract(x, randomPoints(elev, 20000)) %>% data.frame)) %>% do.call(rbind, .)
```


# Model Fitting

```{r cache = T}
tmean.gam <- gam(tmean.obs ~ s(t2m, bs = 'cr') +
             s(msl, bs = 'cr') +
             s(elev, bs = 'cr'),
           method = 'REML', data = cal.data)

p.occur.gam <- gam(factor(p.obs >= .1) ~ s(t2m) + s(cp), 
                   family = binomial, method = 'REML', data = cal.data)

prcp.gam <- bam(p.obs ~ s(msl, bs = 'cr') +
              s(tcw, bs = 'cr') +
              s(lsp, bs = 'cr') +
              s(cp, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(dco, bs = 'cr'),
            family = Gamma(link = 'log'), method = 'REML', 
            data = cal.data[cal.data$p.obs >= .1, ])
```




# Predictions
## Ensemble Generation
Write a function to import, process, and generate a monthly average enemble.
```{r}
getEns <- function(period, variable){
  var.dir <- paste0('~/gdrive/Data/PMIP3 Data/', period, '/', variable)
  files.in <- list.files(var.dir, full.names = T)
  
  sapply(files.in, function(x){
    brick(x) %>% rotate %>% projectRaster(elev)
  }) %>% brick %>% stackApply(indices = 1:12, fun = mean)
}
```

### Mid Holocene
Use this function to import all the necessary variables. Generate a single prediction set
```{r}
t2m <- getEns('MH', 'tas')
msl <- getEns('MH', 'psl')
cp <- getEns('MH', 'prc' ) %>% multiply_by(86.4) 
lsp <- (getEns('MH', 'pr') %>% multiply_by(86.4))  - cp 
tcw <- getEns('MH', 'clivi') + getEns('MH', 'clwvi') + getEns('MH', 'prw')

pred.vars.mh <- sapply(1:12, function(x){
  brick(t2m[[x]], msl[[x]], tcw[[x]], cp[[x]], lsp[[x]], elev, dco) %>%
    setNames(c('t2m', 'msl', 'tcw', 'cp', 'lsp', 'elev', 'dco'))
}) 
```

Make predictions for the Mid Holocene.
```{r}
tmean.mh <- sapply(1:12, function(x){
    predict(pred.vars.mh[[x]], tmeann.gam)
  }) %>% brick
  
prec.occur.mh <- sapply(1:12, function(x){
    predict(pred.vars.mh[[x]], p.occur.gam, type = 'response')
  }) %>% brick %>% is_weakly_greater_than(.5)
  
prec.mh <- sapply(1:12, function(x){
    predict(pred.vars.mh[[x]], prcp.gam, type = 'response')
  }) %>% brick %>% mask(prec.occur.mh, maskvalue = 0, updatevalue = 0)
```

### LGM
Repeat for the LGM
```{r}
t2m <- getEns('LGM', 'tas')
msl <- getEns('LGM', 'psl')
cp <- getEns('LGM', 'prc' ) %>% multiply_by(86.4) 
lsp <- (getEns('LGM', 'pr') %>% multiply_by(86.4))  - cp 
tcw <- getEns('LGM', 'clivi') + getEns('LGM', 'clwvi') + getEns('LGM', 'prw')

pred.vars.lgm <- sapply(1:12, function(x){
  brick(t2m[[x]], msl[[x]], tcw[[x]], cp[[x]], lsp[[x]], elev, dco) %>%
    setNames(c('t2m', 'msl', 'tcw', 'cp', 'lsp', 'elev', 'dco'))
}) 
```

```{r}
tmean.lgm <- sapply(1:12, function(x){
    predict(pred.vars.lgm[[x]], tmean.gam)
  }) %>% brick
  
prec.occur.lgm <- sapply(1:12, function(x){
    predict(pred.vars.lgm[[x]], p.occur.gam, type = 'response')
  }) %>% brick %>% is_weakly_greater_than(.5)
  
prec.lgm <- sapply(1:12, function(x){
    predict(pred.vars.lgm[[x]], prcp.gam, type = 'response')
  }) %>% brick %>% mask(prec.occur.lgm, maskvalue = 0, updatevalue = 0)
```


Now let's import previously-downscaled ensemble equilibrium simulations of the Last Glacial Maximum and Mid Holocene, to estimate how the spatial patterns of climate variability have changed over time, and to test for consistency with the transient TraCE simulation.

First import the downscaled data.
```{r}
bbox <- extent(c(-10, 20, 35, 47))

lgm.prc <- prec.lgm %>% crop(bbox)
mh.prc <- prec.mh %>% crop(bbox)

lgm.tmp <- tmean.lgm %>% crop(bbox)
mh.tmp <- tmean.mh %>% crop(bbox)
```


Calculate changes in seasonal precipitation and temperature.
```{r}
bySeason <- function(x, season, var){
  if(season == 'djf') {ids <- c(1,2,12)}
  if(season == 'jja') {ids <- c(6,7,8)}
  
  if(var == 'tmp') return(mean(x[[ids]]))
  if(var == 'prc') return(sum(x[[ids]]))
}

prc.change.map <- brick(c(
  bySeason(mh.prc, 'djf', 'prc') - bySeason(lgm.prc, 'djf', 'prc'),
  bySeason(mh.prc, 'jja', 'prc') - bySeason(lgm.prc, 'jja', 'prc')))
prc.change.map[prc.change.map < -200] <- -200 # so the plot isn't washed out by large precip values


prc.change.map.percent <- brick(c(
  (bySeason(mh.prc, 'djf', 'prc') - bySeason(lgm.prc, 'djf', 'prc')) * 100 / bySeason(lgm.prc, 'djf', 'prc'), 
  (bySeason(mh.prc, 'jja', 'prc') - bySeason(lgm.prc, 'jja', 'prc')) * 100 / bySeason(lgm.prc, 'jja', 'prc')))

tmp.change.map <- brick(c(
  bySeason(mh.tmp, 'djf', 'tmp') - bySeason(lgm.tmp, 'djf', 'tmp'),
  bySeason(mh.tmp, 'jja', 'tmp') - bySeason(lgm.tmp, 'jja', 'tmp')))
```

Plot the results
```{r}
levelplot(prc.change.map.percent, margin = F, names.attr = c('Winter', 'Summer'),
          main = 'Precipitation Change (%)\n LGM to Mid Holocene',
          par.settings = PuOrTheme(), 
          at = seq(-100,100,10))

# note in the next plot all values less than -200 have been turned into -200
levelplot(prc.change.map, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Precipitation Change (mm) \n LGM to Mid Holocene',
          par.settings = PuOrTheme(), 
          at = seq(-200,200,20))    

levelplot(tmp.change.map, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Temperature Change (°C)\n LGM to Mid Holocene',
          par.settings = BuRdTheme(),
          at = seq(-20,20,2))
```

Now we can calculate changes in seasonality. For temperature, this is just the standard deviation of all 12 monthly averages. For precipitation, we will use the coefficient of variation.
```{r}
tmp.seasonality <- calc(mh.tmp, sd) - calc(lgm.tmp, sd)
prc.seasonality <- cv(mh.prc) - cv(lgm.prc)
```
Plot the results.
```{r}
levelplot(tmp.seasonality, margin = F, 
          main = 'Change in temperature seasonality (SD)\n LGM to Mid Holocene', 
          par.settings = BuRdTheme(), 
          at = seq(-4, 4, .4))

levelplot(prc.seasonality, margin = F, 
          main = 'Change in precipitation seasonality (CV)\n LGM to Mid Holocene', 
          par.settings = PuOrTheme(), 
          at = seq(-50, 50, 5))
```

What about changes in spatial hetergeneity?
```{r}
wts <- matrix(c(0,0,1,0,0,0,1,1,1,0,1,1,1,1,1,0,1,1,1,0,0,0,1,0,0), nrow = 5)

tmp.hetero <- brick(c(
  bySeason(mh.tmp, 'djf', 'tmp') %>%
    focal(w = wts, sd, na.rm = T) %>%
    subtract(
      bySeason(lgm.tmp, 'djf', 'tmp') %>% 
        focal(w = wts, sd, na.rm = T)),
  bySeason(mh.tmp, 'jja', 'tmp') %>%
    focal(w = wts, sd, na.rm = T) %>%
    subtract(
      bySeason(lgm.tmp, 'jja', 'tmp') %>% 
        focal(w = wts, sd, na.rm = T)))) %>%
  mask(mh.tmp[[1]]) # clip buffer added by window
  
levelplot(tmp.hetero, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Temperature heterogeneity (SD in 25km radius) change\n LGM to Mid Holocene',
          par.settings = BuRdTheme(), at = seq(-10, 10, 1))
```
Same for precipitaiton.

```{r}
prc.hetero.sd <- brick(c(
  bySeason(mh.prc, 'djf', 'prc') %>%
    focal(w = wts, sd, na.rm = T) %>%
    subtract(
      bySeason(lgm.prc, 'djf', 'prc') %>% 
        focal(w = wts, sd, na.rm = T)),
  bySeason(mh.prc, 'jja', 'prc') %>%
    focal(w = wts, sd, na.rm = T) %>%
    subtract(
      bySeason(lgm.prc, 'jja', 'prc') %>% 
        focal(w = wts, sd, na.rm = T)))) %>%
  mask(mh.prc[[1]]) # clip buffer added by window

#levelplot(prc.hetero.sd, margin = F, names.attr = c('Winter', 'Summer'), 
#          main = 'Precipitation heterogeneity (SD in 25km radius) change\n LGM to Mid Holocene',
#          par.settings = BuRdTheme(), at = seq(-500, 500, 50))

#capped at -75
prc.hetero.sd[prc.hetero.sd < -75] <- -75
levelplot(prc.hetero.sd, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Precipitation heterogeneity (SD in 25km radius) change\n LGM to Mid Holocene',
          par.settings = BuRdTheme(), at = seq(-75, 75, 7.5))

# capped at -60
prc.hetero.sd[prc.hetero.sd < -60] <- -60
levelplot(prc.hetero.sd, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Precipitation heterogeneity (SD in 25km radius) change\n LGM to Mid Holocene',
          par.settings = BuRdTheme(), at = seq(-60, 60, 6))
```

```{r}
## now cv

prc.hetero.cv <- brick(c(
  bySeason(mh.prc, 'djf', 'prc') %>%
    focal(w = wts, cv, na.rm = T) %>%
    subtract(
      bySeason(lgm.prc, 'djf', 'prc') %>% 
        focal(w = wts, cv, na.rm = T)),
  bySeason(mh.prc, 'jja', 'prc') %>%
    focal(w = wts, cv, na.rm = T) %>%
    subtract(
      bySeason(lgm.prc, 'jja', 'prc') %>% 
        focal(w = wts, cv, na.rm = T)))) %>%
  mask(mh.prc[[1]]) # clip buffer added by window
  
#levelplot(prc.hetero.cv, margin = F, names.attr = c('Winter', 'Summer'), 
#          main = 'Precipitation heterogeneity (CV in 25km radius) change\n LGM to Mid Holocene',
#          par.settings = BuRdTheme(), at = seq(-2.2, 2.2, .22))

# now cap at +/- .25

prc.hetero.cv[prc.hetero.cv > .25] <- .25
prc.hetero.cv[prc.hetero.cv < -.25] <- -.25
levelplot(prc.hetero.cv, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Precipitation heterogeneity (CV in 25km radius) change\n LGM to Mid Holocene',
          par.settings = BuRdTheme(), at = seq(-.25, .25, .025))
```
