---
title: 'Supplementary Information: Paleoclimate Modeling'
author: "Nicolas Gauthier"
date: 'Last Updated: `r Sys.Date()`'
link-citations: yes
mainfont: Times New Roman
monofont: Ubuntu Mono
geometry: margin=1in
fontsize: 12pt
header-includes:
- \usepackage{marginnote}
output:
  pdf_document:
    fig_height: 5
    fig_width: 10
    highlight: pygments
    latex_engine: lualatex
---

```{r}
library(raster)
library(magrittr)
library(dismo)
library(mgcv)
```


# Data Preprocessing

## Observations
```{r}
bbox <- extent(c(-10, 45, 30, 50))
  
tmean.obs <- list.files('~/gdrive/Data/MOD11C3v5.0-CHIRPSv2.0_MONTHLY_03m/meantemp', full.names = T) %>% 
    stack %>%
    set_names(month.name) %>%
    crop(bbox)

p.obs <- list.files('~/gdrive/Data/MOD11C3v5.0-CHIRPSv2.0_MONTHLY_03m/precip', full.names = T) %>% 
    stack %>%
    set_names(month.name) %>%
    crop(bbox)
```


## Topographic Variables
```{r}
elev <- raster('~/gdrive/Data/SRTM_1km.tif') %>% projectRaster(p.obs) %>% mask(p.obs)
```


```{r}
dco <- elev %>%
  reclassify(c(-Inf, Inf, NA, NA, NA, 1)) %>% # reverse NA and non-NA cells
  distance(doEdge = T) %>% # calculate the distances
  mask(elev) %>% # mask out ocean cells  
  divide_by(1000) # convert to km
```



## Reanalysis
Import and preprocess of ECMWF-interim reanalysis data, monthly means of daily means, 1979-2010
```{r eval = FALSE}
processECMWF <- function(file, var){
  brick(paste0('~/gdrive/Data/', file), varname = var) %>% 
      stackApply(indices = 1:12, fun = mean) %>%
      rotate %>%
      set_names(month.name) %>% projectRaster(eu.p) %>% mask(eu.p)
}

tcw <- processECMWF('ecmwf_surface.nc', 'tcw')
msl <- processECMWF('ecmwf_surface.nc', 'msl')
t2m <- processECMWF('ecmwf_surface.nc', 't2m')
lsp <- processECMWF('ECMWF Precip.nc', 'lsp')
cp <- processECMWF('ECMWF Precip.nc', 'cp')
```

```{r, include=FALSE}
# load preloaded files at first for convience, should change
msl <- brick('~/gdrive/Data/ECMWF/eu/msl_ecmwf_eu.tif')
t2m <- brick('~/gdrive/Data/ECMWF/eu/t2m_ecmwf_eu.tif')
tcw <- brick('~/gdrive/Data/ECMWF/eu/tcw_ecmwf_eu.tif')
cp <- brick('~/gdrive/Data/ECMWF/eu/cp_ecmwf_eu.tif')
lsp <- brick('~/gdrive/Data/ECMWF/eu/lsp_ecmwf_eu.tif')
```


Put all the predictor and response variables together, month by month.
```{r}
cal.vars <- sapply(1:12, function(x){ 
  brick(tmean.obs[[x]], p.obs[[x]], msl[[x]], t2m[[x]], tcw[[x]], lsp[[x]], cp[[x]], elev, dco) %>%
    setNames(c('tmean.obs', 'p.obs', 'msl', 't2m', 'tcw', 'lsp', 'cp','elev','dco'))
})
```

Sample the variables at random points, weighting for latitude
```{r}
cal.data <- lapply(cal.vars, function(x) (raster::extract(x, randomPoints(elev, 20000)) %>% data.frame)) %>% do.call(rbind, .)
```


# Model Fitting

```{r}
tmean.gam <- gam(tmean.obs ~ s(t2m, bs = 'cr') +
             s(msl, bs = 'cr') +
             s(elev, bs = 'cr'),
           method = 'REML', data = cal.data)

p.occur.gam <- gam(factor(p.obs >= .1) ~ s(t2m) + s(cp), 
                   family = binomial, method = 'REML', data = cal.data)

prcp.gam <- bam(p.obs ~ s(msl, bs = 'cr') +
              s(tcw, bs = 'cr') +
              s(lsp, bs = 'cr') +
              s(cp, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(dco, bs = 'cr'),
            family = Gamma(link = 'log'), method = 'REML', 
            data = cal.data[cal.data$p.obs >= .1, ])
```




# Predictions
## Ensemble Generation
Write a function to import, process, and generate a monthly average enemble.
```{r}
getEns <- function(period, variable){
  var.dir <- paste0('~/gdrive/Data/PMIP3 Data/', period, '/', variable)
  files.in <- list.files(var.dir, full.names = T)
  
  sapply(files.in, function(x){
    brick(x) %>% rotate %>% projectRaster(elev)
  }) %>% brick %>% stackApply(indices = 1:12, fun = mean)
}
```

### Mid Holocene
Use this function to import all the necessary variables. Generate a single prediction set
```{r}
t2m <- getEns('MH', 'tas')
msl <- getEns('MH', 'psl')
cp <- getEns('MH', 'prc' ) %>% multiply_by(86.4) 
lsp <- (getEns('MH', 'pr') %>% multiply_by(86.4))  - cp 
tcw <- getEns('MH', 'clivi') + getEns('MH', 'clwvi') + getEns('MH', 'prw')

pred.vars.mh <- sapply(1:12, function(x){
  brick(t2m[[x]], msl[[x]], tcw[[x]], cp[[x]], lsp[[x]], elev, dco) %>%
    setNames(c('t2m', 'msl', 'tcw', 'cp', 'lsp', 'elev', 'dco'))
}) 
```

Make predictions for the Mid Holocene.
```{r}
tmean.mh <- sapply(1:12, function(x){
    predict(pred.vars.mh[[x]], tmeann.gam)
  }) %>% brick
  
prec.occur.mh <- sapply(1:12, function(x){
    predict(pred.vars.mh[[x]], p.occur.gam, type = 'response')
  }) %>% brick %>% is_weakly_greater_than(.5)
  
prec.mh <- sapply(1:12, function(x){
    predict(pred.vars.mh[[x]], prcp.gam, type = 'response')
  }) %>% brick %>% mask(prec.occur.mh, maskvalue = 0, updatevalue = 0)
```

### LGM
Repeat for the LGM
```{r}
t2m <- getEns('LGM', 'tas')
msl <- getEns('LGM', 'psl')
cp <- getEns('LGM', 'prc' ) %>% multiply_by(86.4) 
lsp <- (getEns('LGM', 'pr') %>% multiply_by(86.4))  - cp 
tcw <- getEns('LGM', 'clivi') + getEns('LGM', 'clwvi') + getEns('LGM', 'prw')

pred.vars.lgm <- sapply(1:12, function(x){
  brick(t2m[[x]], msl[[x]], tcw[[x]], cp[[x]], lsp[[x]], elev, dco) %>%
    setNames(c('t2m', 'msl', 'tcw', 'cp', 'lsp', 'elev', 'dco'))
}) 
```

```{r}
tmean.lgm <- sapply(1:12, function(x){
    predict(pred.vars.lgm[[x]], tmean.gam)
  }) %>% brick
  
prec.occur.lgm <- sapply(1:12, function(x){
    predict(pred.vars.lgm[[x]], p.occur.gam, type = 'response')
  }) %>% brick %>% is_weakly_greater_than(.5)
  
prec.lgm <- sapply(1:12, function(x){
    predict(pred.vars.lgm[[x]], prcp.gam, type = 'response')
  }) %>% brick %>% mask(prec.occur.lgm, maskvalue = 0, updatevalue = 0)
```




