---
title: "LGM Europe - Downscaled CCSM4"
author: "Nick Gauthier"
date: "January 6, 2016"
output: html_document
---

In this notebook, we estimate LGM monthly mean precipitation and minimum and maximum precipitation in the western Mediterranean by statistically a downscaling General Circulation Model (GCM) paleoclimate simulation. We first calibrate a Genralized Additive Model (GAM) using observed WorldClim climatologies, DEM-derived topographic variables, and simulations of present day climates using NCAR's CESM/CCSM4 model. Then we apply this model to CCSM4 simulations of the LGM PMIP3 experiments.
 
 ##Preparation
 ###Setup
 
 First load the necessary R packages.

``` {r message= F, warning = F}
library(raster)     # functions for managing and processing raster data
library(mgcv)       # functions to fit and analyze GAMs
library(dismo)      # functions to sample points weighted by latitude
library(magrittr)   # piping functions for code readability
library(rasterVis)  # functions for visualizing raster maps
library(ggplot2)    # functions for plotting
library(reshape2)
```

### Observed Climate

Import the observed present-day climatologies which we'll use to calibrate the model. We use WorldClim data at 5min resolution here, which can be easily downloaded using the *getData()* function in the *raster* package. 

```{r, message=F}
get.wc <- function(var){
     raster::getData('worldclim', var=var, res = 5, download = T) %>% 
     crop(extent(-20, 40, 30, 58)) %>%     
     set_names(month.name)
}

obs.prc <- get.wc('prec')
obs.tmx <- get.wc('tmax') %>% divide_by(10) # raw data in degrees Celsius * 10
obs.tmn <- get.wc('tmin') %>% divide_by(10) 
```

```{r echo = F}
levelplot(obs.prc[[c(1, 7)]], par.settings = rasterTheme(region=brewer.pal('Blues', n=9)))

levelplot(brick(c(obs.tmn[[c(1,7)]], obs.tmx[[c(1,7)]])), par.settings = BuRdTheme)

ggplot(melt(values(obs.prc)), aes(x=Var2, y=value, group=Var1)) + 
     geom_line(alpha = .01) +
     theme_classic()

ggplot(melt(values(obs.tmx)), aes(x=Var2, y=value, group=Var1)) + 
     geom_line(alpha = .01) +
     theme_classic()
```


### GCM Predictors
```{r}
import.gcm <- function(dir, var, sim = 'lgm', level = 1){
     gcm.in <- brick(dir, var = var, level = level) 
     
     if(sim == 'control'){gcm.in <- gcm.in[[1200:1799]]}
     if(sim == 'lgm'){gcm.in <- gcm.in[[4213:4812]]}
     
     gcm.in <- stackApply(gcm.in, indices = 1:12, fun = mean)
     extent(gcm.in) <- extent(0, 360, -90, 90)
     
     gcm.in %>%    
     rotate %>%
     projectRaster(obs.prc) %>%
     mask(obs.prc)
}

control.prc <- import.gcm('GCM/b40.20th.track1.1deg.005.cam2.h0.PRECT.185001-200512.nc', 
                            var = "PRECT", 
                            sim = 'control') %>% multiply_by(2629743830)

control.tmx <- import.gcm('GCM/b40.20th.track1.1deg.005.cam2.h0.TREFMXAV.185001-200512.nc', 
                            var = "TREFMXAV", 
                            sim = 'control') %>% subtract(273.15)

control.tmn <- import.gcm('GCM/b40.20th.track1.1deg.005.cam2.h0.TREFMNAV.185001-200512.nc', 
                            var = "TREFMNAV", 
                            sim = 'control') %>% subtract(273.15)

control.psl <- import.gcm('GCM/b40.20th.track1.1deg.005.cam2.h0.PSL.185001-200512.nc', 
                            var = "PSL", 
                            sim = 'control') %>% divide_by(1000)

control.q <- import.gcm('GCM/b40.20th.track1.1deg.005.cam2.h0.QREFHT.185001-200512.nc', 
                           var = "QREFHT", 
                           sim = 'control')

control.u <- import.gcm('GCM/b40.20th.track1.1deg.005.cam2.h0.U.185001-200512.nc', 
                          var = "U", 
                          sim = 'control',
                          level = 26)

control.v <- import.gcm('GCM/b40.20th.track1.1deg.005.cam2.h0.V.185001-200512.nc', 
                          var = "V", 
                          sim = 'control',
                          level = 26)

control.z <- import.gcm('GCM/b40.20th.track1.1deg.005.cam2.h0.Z3.185001-200512.nc', 
                          var = "Z3", 
                          sim = 'control',
                          level = 23) # ~850 geopotential
```

### Topographic Predictors
Generate some topographic predictors from a DEM.

Start with importing a DEM, then calculate aspect and slope maps using the **terrain** function.
```{r}
elev <- raster('alt_5m_bil/alt.bil') %>% projectRaster(obs.prc)
aspect <- elev %>% terrain(opt='aspect', unit = 'degrees')
slope <- elev %>% terrain(opt = 'slope', unit = 'degrees')
```

```{r echo = F}
levelplot(elev, par.settings = GrTheme, margin = F)
levelplot(slope, par.settings = GrTheme, margin = F)
levelplot(aspect, par.settings = GrTheme, margin = F)
```

Calculate the euclidean distance to the ocean, accounting for latitude effects.
```{r}
dco <- raster('alt_5m_bil/alt.bil') %>% # import WorldClim 5m dem
    crop(extent(-20, 45, 30, 65)) %>%  # start with a wider region to get accurate distances
    reclassify(c(-Inf, Inf, NA, NA, NA, 1)) %>% # reverse NA and non-NA cells
    distance(doEdge = T) %>% # calculate the distances
    projectRaster(elev) %>%
    mask(elev) %>% # mask out ocean cells  
    divide_by(1000) # convert to km
```

```{r echo = F}
levelplot(dco, margin = F)
```



```{r}
streamplot(stack(control.u[[1]], control.v[[1]]), isField = 'dXY')
streamplot(stack(control.u[[7]], control.v[[7]]), isField = 'dXY')

dir <- (270 - overlay(control.v,  control.u, fun = atan2) * (180 / pi)) %% 360
vel <- sqrt(control.u ^ 2 + control.v ^ 2)

delta <- abs(dir - aspect) #distance between wind angle and slope orientation
values(delta) <- ifelse(values(delta) > 180, 360 - values(delta), values(delta))
vg <- sin(slope * 2 * pi / 180) * cos(delta * pi / 180) * vel * .5
```

```{r echo=F}
plot(vg[[1]])
```

We can expect the model errors to vary by month and by location (i.e. spatially and temporally autocorrelated residuals). Let's include these variables in the model fitting.
```{r}
month <- obs.prc %>% setValues(c(as.factor(rep(1:12, each = 241920)))) # each should = ncell(vg)

lat <- elev %>%
     coordinates %>%
     extract( ,2) %>% 
     setValues(elev, .)

lon <- elev %>%
     coordinates %>%
     extract( ,1) %>% 
     setValues(elev, .)
```

Put everything together.
```{r}
cal.vars <- sapply(1:12, function(x){ 
  brick(obs.tmx[[x]], obs.tmn[[x]], obs.prc[[x]], control.tmx[[x]], 
    control.tmn[[x]], control.q[[x]], control.prc[[x]], control.psl[[x]], 
    control.z[[x]], elev, dco, vg[[x]], month[[x]], lat, lon) %>%
  setNames(c('obs.tmx', 'obs.tmn', 'obs.prc','TREFMXAV', 'TREFMNAV', 'Q', 'PRC', 
    'PSL', 'Z3', 'elev','dco', 'vg', 'month', 'lat','lon'))
})

rm(control.tmx, control.tmn, control.q,control.prc,control.psl,control.z,control.v,control.u,vg,dir,vel,delta)
```

Sample the variables at random points, weighting for area distortions due to latitude
```{r}
cal.data <- lapply(cal.vars, function(x) (raster::extract(x, randomPoints(elev, 20000)) %>% data.frame)) %>% 
  do.call(rbind, .)
```

## Fit the model
let's start with temperature
```{r}
fit.tmn <- gam(obs.tmn ~ s(TREFMNAV, bs = 'cr') +
                         s(Z3, bs = 'cr') +
                         s(elev, bs = 'cr') +
                         s(dco, bs = 'cr') +
                         s(month, bs = 're'), 
               method = 'REML', data = cal.data)


fit.tmn
summary(fit.tmn)
gam.check(fit.tmn)
plot(fit.tmn, shade=T, seWithMean = T, pages = 1)
```

```{r}
fit.tmx <- gam(obs.tmx ~ s(TREFMXAV, bs = 'cr') +
                         s(Z3, bs = 'cr') +
                         s(elev, bs = 'cr') +
                         s(dco, bs = 'cr') +
                         s(month, bs = 're'),
                method = 'REML', data = cal.data)

fit.tmx
summary(fit.tmx)
gam.check(fit.tmn)
plot(fit.tmx, shade = T, seWithMean = T, pages = 1)
```

```{r}
fit.prc.occur <- gam(factor(obs.prc >= 1) ~ s(PRC),
                    family = binomial, method = 'REML', data = cal.data)

summary(fit.prc.occur)
gam.check(fit.prc.occur)
plot(fit.prc.occur, seWithMean = T, shade = T, pages = 1)
```

```{r}
fit.prc <- bam(obs.prc ~ s(PRC, bs = 'cr') +
                         s(PSL, bs = 'cr') +
                         s(Q, bs = 'cr') + 
                         s(vg, bs = 'cr') +
                         s(Z3, bs = 'cr') +
                         s(elev, bs = 'cr') +
                         s(dco, bs = 'cr') +
                         s(month, bs = 're'),
               family = Gamma(link = 'log'), method = 'REML', data = cal.data[cal.data$obs.prc>=1,])

fit.prc
summary(fit.prc)
gam.check(fit.prc)
plot(fit.prc, shade = T, seWithMean = T, pages = 1)
```


# Paleo
```{r}
lgm.tmx <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.TREFMXAV.149901-189912.nc', 
                         var = "TREFMXAV") %>% subtract(273.15)

lgm.tmn <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.TREFMNAV.149901-189912.nc', 
                         var = "TREFMNAV") %>% subtract(273.15)

lgm.q <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.QREFHT.149901-189912.nc', 
                           var = "QREFHT")

lgm.psl <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.PSL.149901-189912.nc', 
                         var = "PSL") %>% divide_by(1000)

lgm.prcc <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.PRECC.149901-189912.nc', 
                         var = "PRECC") %>% multiply_by(2629743830)

lgm.prcl <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.PRECL.149901-189912.nc', 
                         var = "PRECL") %>% multiply_by(2629743830)

lgm.prc <- lgm.prcl + lgm.prcc

lgm.u <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.U.149901-189912.nc', 
                          var = "U",
                          level = 26)

lgm.v <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.V.149901-189912.nc', 
                          var = "V",
                          level = 26)

lgm.z <- import.gcm('GCM/b40.lgm21ka.1deg.003.cam2.h0.Z3.149901-189912.nc', 
                         var = "Z3", level = 23)

streamplot(stack(lgm.u[[1]], lgm.v[[1]]), isField = 'dXY')
streamplot(stack(lgm.u[[7]], lgm.v[[7]]), isField = 'dXY')

lgm.dir <- (270 - overlay(lgm.v,  lgm.u, fun = atan2) * (180 / pi)) %% 360
lgm.vel <- sqrt(lgm.u ^ 2 + lgm.v ^ 2)
lgm.delta <- abs(lgm.dir - aspect)
values(lgm.delta) <- ifelse(values(lgm.delta) > 180, 360 - values(lgm.delta), values(lgm.delta))
lgm.vg <- sin(slope * 2 * pi / 180) * cos(lgm.delta * pi / 180) * lgm.vel * .5

pred.vars <- sapply(1:12, function(x){ 
  brick(lgm.tmx[[x]], lgm.tmn[[x]], lgm.prc[[x]], lgm.psl[[x]], lgm.q[[x]], lgm.z[[x]], elev, dco, lgm.vg[[x]], month[[x]]) %>% 
    crop(extent(-5, 15, 37, 47)) %>%
    setNames(c('TREFMXAV', 'TREFMNAV', 'PRC', 'PSL', 'Q','Z3', 'elev','dco', 'vg','month'))
})

rm(lgm.tmn,lgm.tmx,lgm.z,lgm.q,lgm.v,lgm.u,lgm.prcc,lgm.prcl,lgm.vg,lgm.prc,lgm.dir,lgm.psl,lgm.vel,lgm.delta)
```

```{r}
tmn.recon <- lapply(1:12, function(x){predict(pred.vars[[x]], fit.tmn, type = 'response')}) %>% brick
tmx.recon <- lapply(1:12, function(x){predict(pred.vars[[x]], fit.tmx, type = 'response')}) %>% brick
prc.occur.recon <- lapply(1:12, function(x){predict(pred.vars[[x]], fit.prc.occur, type = 'response')}) %>% brick
prc.recon <- lapply(1:12, function(x){predict(pred.vars[[x]], fit.prc, type = 'response')}) %>% brick
prc.recon <- mask(prc.recon, prc.occur.recon>.7, maskvalue=0, update.value=0)
```

```{r echo = F}
levelplot(prc.occur.recon)
levelplot(prc.recon - obs.prc, par.settings = RdBuTheme, at = seq(-300,300,20))
levelplot(tmn.recon - obs.tmn, par.settings = BuRdTheme, at = seq(-100,100,10))
levelplot(tmx.recon - obs.tmx, par.settings = BuRdTheme, at = seq(-100,100,10))
```

```{r eval = F, echo = F}
writeRaster(tmn.recon, 'tmn_lgm.tif', format = 'GTiff', options="INTERLEAVE=BAND", overwrite = T)
writeRaster(tmx.recon, 'tmx_lgm.tif', format = 'GTiff', options="INTERLEAVE=BAND", overwrite = T)
writeRaster(prc.recon, 'prc_lgm.tif', format = 'GTiff', options="INTERLEAVE=BAND", overwrite = T)
```
