---
title: "West Mediterranean during the Last Deglaciation"
author: "Nick Gauthier"
date: '`r Sys.Date()`'
output:
  tufte::tufte_html:
    keep_md: yes
    tufte_features:
    - fonts
    - italics
  tufte::tufte_handout:
    citation_package: natbib
    highlight: pygments
    keep_tex: yes
link-citations: yes
mainfont: Palatino
monofont: Bera Mono
geometry: margin=1in
sansfont: Helvetica
subtitle: A proxy -- model intercomparison
fontsize: 11pt
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Introduction

We'll be comparing paleoclimate model estimates of temperature and precipitation over three points in the west Mediterranean to global paleoclimate proxies.

## Setup
Load all the packages we'll need for this analysis. 

`r margin_note("You'll need to have the netCDF libraries already installed on your system for ncdf4 to work.")`
```{r libraries, message=F}
library(ncdf4) # import GCM data
library(rgdal) # read GCM data
library(raster) # process GCM data
library(rasterVis) # plotting GCM data
library(tidyverse) # data management and plotting
library(magrittr) # pipes for code readability
library(EMD) # calculate trends in the data
```


# Climate Model

## Sample Locations
```{r map, echo = F, fig.margin = T, message = F, cache = F, fig.cap = 'Locations of 3 sample points.'}
library(ggmap)

samp.pts <- matrix(c(1,40, 4,42, 14,46), ncol = 2, byrow = T) %>% as_data_frame

get_map(location = c(left = -10, bottom = 35, right = 20, top = 50), 
        maptype = 'terrain-background', source = 'stamen') %>% 
  ggmap + labs(x = 'Longitude', y = 'Latitude') + geom_point(data = samp.pts, aes(x = V1, y = V2), size = 3, color = 'red')
```

Create a matrix with the coordinates for the three locations of interest in the west Mediterranean. We'll be focusing on large grid cell averages, so the points do not have to be directly over land.
```{r samppoints}
samp.pts <- matrix(c(1, 40, 4, 42, 14, 46), 
                   ncol = 2, byrow = T)
```

## TraCE-21k

```{r trace_map, echo=FALSE, fig.show = 'hold', fig.cap = 'TraCE21-k global precipitation and temperature', fig.margin=TRUE, message=F, cache = F}
brick('trace.01-36.22000BP.cam2.PRECT.22000BP_decavg_400BCE.nc') %>%
  extract2(1) %>%
  multiply_by(3.154e+10) %>%
  levelplot(margin = F, par.settings = PuOrTheme())

brick('trace.01-36.22000BP.cam2.TREFHT.22000BP_decavg_400BCE.nc') %>%
  extract2(1) %>%
  subtract(273.15)%>%
  levelplot(margin = F, par.settings = BuRdTheme())
```

First, import data from the TraCE-21k paleoclimate simulation. Then extract temperature and precipitation values at three locations in the west Mediterranean. Use the *brick* function from **raster** to import decadal averages from the simulation. Put the coordinates for the three locations in a matrix, and use that matrix to and **raster's** *extract* function to get the values from the climate model brick. Convert the precipitation values to mm/year and temperature values to degrees Celsius. Finally, name the columns for each region appropriately. 

Now pull all the TraCE data into one data frame, with one row per year, and one column per variable/location combination. First *rbind* the two sets of TraCE data and *transpose* the results, turning the 6 rows into 6 columns. Add a column for the Year (in ka BP), and use to select only the entries earlier than 6,000 BP.

```{r}
trace.dat <- rbind(
  brick('trace.01-36.22000BP.cam2.PRECT.22000BP_decavg_400BCE.nc') %>%
    raster::extract(samp.pts) %>% # extract data at these coordinates
    multiply_by(3.154e+10), # convert to mm/year
  brick('trace.01-36.22000BP.cam2.TREFHT.22000BP_decavg_400BCE.nc') %>%
    raster::extract(samp.pts) %>% 
    subtract(273.15)) %>% # convert from kelvin to C
  t %>% # transpose
  as.data.frame %>%
  set_colnames(c('tmp,Southwest', 'tmp,North Central', 'tmp,Northeast', 
                 'prc,Southwest', 'prc,North Central', 'prc,Northeast')) %>%
  rownames_to_column('Year') %>%
  mutate(Year = as.numeric(substring(Year, 3))) %>%
  filter(Year > 6) # get all the decades up to 6ka BP
```



## Trend Analysis

Let's use the **EMD** package to calculate actual trend lines using the empirical mode decomposition approach.
```{r, echo=FALSE, fig.cap="Empirical mode decomposition", fig.height = 12, fig.margin=TRUE, message=F, cache = F}
emd.test <- emd(trace.dat[,2])

par(mfrow=c(emd.test$nimf+1, 1), mar=c(2,1,2,1))
rangeimf <- range(emd.test$imf)
for(i in 1:emd.test$nimf) {
  plot(1:1600, emd.test$imf[,i], type="l", xlab="", ylab="", ylim=rangeimf,
       main=paste(i, "-th IMF", sep="")); abline(h=0)
}
```

Now organize the temperature and precipitation data to make plotting easier using functions from **tidyr**.
`r margin_note("Replace the variable names to make facet naming easier too.")`
```{r cache = F}
trace.plot <- trace.dat %>% 
  gather(key, value, - Year) %>%
  separate(key, c('Variable', 'Region'), ',') %>%
  mutate(Variable = ifelse(
    Variable == 'tmp', 'Temperature (°C)', 'Precipitation (mm)'))

emd.res <- function(x) emd(x)$residue
trace.emd <- trace.dat %>%
  mutate_at(vars(-Year), emd.res) %>%
  gather(key, value, - Year) %>%
  separate(key, c('Variable', 'Region'), ',') %>%
  mutate(Variable = ifelse(
    Variable == 'tmp', 'Temperature (°C)', 'Precipitation (mm)'))
```

## Plotting
Plot everything with **ggplot2**.
```{r bwplot, echo=FALSE, fig.cap="Black and white version", fig.width = 16.9, fig.height = 8.61, fig.margin=TRUE, message=F, cache = F}
ggplot(data = trace.plot, aes(x = Year, y = value)) +
  facet_grid(Variable ~ Region, switch = 'y', scale = 'free_y') +
  geom_vline(xintercept = c(22, 19, 14, 10, 6), lty = 2) +
  geom_point(color = 'grey', alpha = .3) +
  geom_line(data = trace.emd, size = 1.2, color = "black", alpha = .8) +
  scale_x_reverse(breaks = seq(6,22,4)) +
  guides(color = "none") +
  labs(x = '1,000 Years BP', y = '') +
  theme_bw(base_size = 20) +
  theme(strip.background  = element_blank())
```

```{r mainplot, fig.fullwidth = TRUE, fig.width = 16.9, fig.height = 8.61}
ggplot(data = trace.plot, aes(x = Year, y = value)) +
  facet_grid(Variable ~ Region, switch = 'y', scale = 'free_y') +
  geom_vline(xintercept = c(22, 19, 14, 10, 6), lty = 2) +
  geom_point(aes(color = Variable), alpha = .3) +
  geom_line(data = trace.emd, size = 1.2, color = "black", alpha = .8) +
  scale_x_reverse(breaks = seq(6,22,4)) +
  labs(x = '1,000 Years BP', y = '') +
  guides(color = "none") +
  theme_bw(base_size = 20) +
  theme(strip.background  = element_blank())
```

## Spatial patterns

Now let's import previously-downscaled ensemble equilibrium simulations of the Last Glacial Maximum and Mid Holocene, to estimate how the spatial patterns of climate variability have changed over time, and to test for consistency with the transient TraCE simulation.

First import the downscaled data.
```{r}
bbox <- extent(c(-10, 20, 35, 47))

lgm.prc <- brick('Downscaled/ensemble_prc_lgm.tif') %>%
  crop(bbox)
mh.prc <- brick('Downscaled/ensemble_prc_mh6k.tif') %>%
  crop(bbox)

lgm.tmp <- brick('Downscaled/ensemble_tmn_lgm.tif') %>%
  crop(bbox) %>%
  add(brick('Downscaled/ensemble_tmx_lgm.tif') %>% crop(bbox)) %>%
  divide_by(2)
mh.tmp <- brick('Downscaled/ensemble_tmn_mh6k.tif') %>%
  crop(bbox) %>%
  add(brick('Downscaled/ensemble_tmx_mh6k.tif') %>% crop(bbox)) %>%
  divide_by(2)
```

Calculate changes in seasonal precipitation and temperature.
```{r}
bySeason <- function(x, season, var){
  if(season == 'djf') {ids <- c(1,2,12)}
  if(season == 'jja') {ids <- c(6,7,8)}
  
  if(var == 'tmp') return(mean(x[[ids]]))
  if(var == 'prc') return(sum(x[[ids]]))
}

prc.change.map <- brick(c(
  (bySeason(mh.prc, 'djf', 'prc') - bySeason(lgm.prc, 'djf', 'prc')) * 100 / bySeason(lgm.prc, 'djf', 'prc'),
  (bySeason(mh.prc, 'jja', 'prc') - bySeason(lgm.prc, 'jja', 'prc')) * 100 / bySeason(lgm.prc, 'jja', 'prc')))

tmp.change.map <- brick(c(
  bySeason(mh.tmp, 'djf', 'tmp') - bySeason(lgm.tmp, 'djf', 'tmp'),
  bySeason(mh.tmp, 'jja', 'tmp') - bySeason(lgm.tmp, 'jja', 'tmp')))
```

Plot the results
```{r}
levelplot(prc.change.map, margin = F, names.attr = c('Winter', 'Summer'),
          main = 'Precipitation Change (%)\n LGM to Mid Holocene',
          par.settings = PuOrTheme(), 
          at = seq(-100,100,10))
levelplot(tmp.change.map, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Temperature Change\n LGM to Mid Holocene',
          par.settings = BuRdTheme(),
          at = seq(-20,20,2))
```

Now we can calculate changes in seasonality. For temperature, this is just the standard deviation of all 12 monthly averages. For precipitation, we will use the coefficient of variation.
```{r}
tmp.seasonality <- calc(mh.tmp, sd) - calc(lgm.tmp, sd)
prc.seasonality <- cv(mh.prc) - cv(lgm.prc)
```
Plot the results.
```{r}
levelplot(tmp.seasonality, margin = F, 
          main = 'Change in temperature seasonality (SD)\n LGM to Mid Holocene', 
          par.settings = BuRdTheme(), 
          at = seq(-4, 4, .4))

levelplot(prc.seasonality, margin = F, 
          main = 'Change in precipitation seasonality (CV)\n LGM to Mid Holocene', 
          par.settings = BuRdTheme(), 
          at = seq(-50, 50, 5))
```

What about changes in spatial hetergeneity?
```{r}
wts <- matrix(c(0,0,1,0,0,0,1,1,1,0,1,1,1,1,1,0,1,1,1,0,0,0,1,0,0), nrow = 5)

tmp.hetero <- brick(c(
  bySeason(mh.tmp, 'djf', 'tmp') %>%
    focal(w = wts, sd, na.rm = T) %>%
    subtract(
      bySeason(lgm.tmp, 'djf', 'tmp') %>% 
        focal(w = wts, sd, na.rm = T)),
  bySeason(mh.tmp, 'jja', 'tmp') %>%
    focal(w = wts, sd, na.rm = T) %>%
    subtract(
      bySeason(lgm.tmp, 'jja', 'tmp') %>% 
        focal(w = wts, sd, na.rm = T)))) %>%
  mask(mh.tmp[[1]]) # clip buffer added by window
  
levelplot(tmp.hetero, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Temperature heterogeneity (SD in 25km radius) change\n LGM to Mid Holocene',
          par.settings = BuRdTheme(), at = seq(-10, 10, 1))
```
Same for precipitaiton.

```{r}
prc.hetero <- brick(c(
  bySeason(mh.prc, 'djf', 'prc') %>%
    focal(w = wts, sd, na.rm = T) %>%
    subtract(
      bySeason(lgm.prc, 'djf', 'prc') %>% 
        focal(w = wts, sd, na.rm = T)),
  bySeason(mh.prc, 'jja', 'prc') %>%
    focal(w = wts, sd, na.rm = T) %>%
    subtract(
      bySeason(lgm.prc, 'jja', 'prc') %>% 
        focal(w = wts, sd, na.rm = T)))) %>%
  mask(mh.prc[[1]]) # clip buffer added by window
  
  
levelplot(prc.hetero, margin = F, names.attr = c('Winter', 'Summer'), 
          main = 'Precipitation heterogeneity (SD in 25km radius) change\n LGM to Mid Holocene',
          par.settings = BuRdTheme(), at = seq(-500, 500, 50))
```

`r margin_note("Compare these with raw gcm outputs to check the added value of SDM")`

# Proxy records

Get ice core data.
```{r message = F}
core.dat <- read_csv('icecores_newdates.csv') %>%
  transmute(years.BP, ngrip =  d18O.NGRIP2.ppt, gisp = d18O.GISP2.ppt) %>%
  filter(years.BP < 24000 & years.BP >=6000)
```
Plot it
```{r}
core.plot <- gather(core.dat, 'core', 'd18O', 2:3)

ggplot(core.plot, aes(x = years.BP, y = d18O, color = core)) +
  geom_line(alpha = .54) +
  geom_smooth() +
  scale_x_reverse(breaks = seq(6000,22000,4000)) +
  theme_minimal()
```

## Detrending
```{r cache = F}
cores <- core.dat %>% na.omit
core.emd.plot <- emd(cores$gisp, cores$years.BP)$residue %>% 
  data_frame(d18O = ., years.BP = cores$years.BP)
```

```{r}
ggplot(core.plot, aes(x = years.BP, y = d18O)) +
  geom_line(aes(color = core), alpha = .54) +
  geom_line(data = core.emd.plot) +
  scale_x_reverse(breaks = seq(6000,22000,4000)) +
  theme_minimal()
```
